String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-04-26 2023-1-19/sections/preliminaries.tex
intro
prelim

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$.  We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}

Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

>>>>

\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$.  We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
%Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
A concatenation of two languages $L_1$ and $L_2$ is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$.
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

### 2023-04-26 2023-1-19/sections/recl.tex
prelim

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length function (abbreviated as RECL). The syntax of RECL is defined by the rules in Table~\ref{tab:syntax}. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$, $x \not \in \regex$, or $t_1\ \op\ t_2$, where $x$ is a string variable, $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ or $x \not \in \regex$ are called \emph{regular membership} constraints and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, alternation $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. 
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{table}[h]
  \centering
  \begin{tabular}{l r}
    $\varphi$ ::= $x\in \regex \mid x \not \in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi $                                             & formulas            \\
    $\regex$ ::= $\emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}}$ & regular expressions \\
    $t$ ::= $n \mid i \mid  |x| \mid t - t \mid t + t$                                                                       & integer terms
  \end{tabular}
  \caption{Syntax of RECL }\label{tab:syntax}
\end{table}
% Semantic

@@

%\subsubsection{Semantic}
We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$. $\pi: I\rightarrow\mathbb{Z}$ is the interpretation of Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \pi\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution $(\eta, \pi)$ exists such as $\eta, \pi\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.
\begin{table}[h]
  \centering
  \begin{tabular}{lcl}
    $\eta,\pi \models \varphi_1\vee \varphi_2$ & $\mathsf{iff}$ & $\eta,\pi \models \varphi_1 \text{ or } \eta,\pi \models \varphi_2$ \\
    $\eta,\pi \models \neg\varphi $            & $\mathsf{iff}$ & $\eta,\pi \not\models \varphi$                                      \\
    $\eta, \pi \models x\in \regex$            & $\mathsf{iff}$ & $\exists w \in \lan(\regex),\eta,\pi \models x = w$                 \\
    $\eta, \pi \models \alpha_1 \leq \alpha_2$ & $\mathsf{iff}$ & $\pi(\alpha_1) \leq \pi(\alpha_2) $                                 \\
  \end{tabular}
  \caption{Semantics}
  \label{tab:semantics}
\end{table}
In addition to the classic regex operators (union, concatenation, closure), we syntactically support intersection, complement, and repetition. As we all know, the classic regular language is closed under intersection and complement \cite{aut_hopcraft}. Furthermore, the operation \emph{repetition} $\regex\{m,n\}$ means repeating regex $\regex$ at least $m$ times and at most $n$ times. It can be syntactically rewritten by concatenation and union:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $\regex$ $k$ times ($m\leq k\leq n$). So the regular language $\lan(\regex)$ defined in Table \ref{tab:syntax} is semantically equal to the classic regular language.

>>>>

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length function (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $i$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$, $x \not \in \regex$, or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ or $x \not \in \regex$ are called \emph{regular membership} constraints and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, alternation $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is is built from constants $n$, variables $i$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid x \not \in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid i \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

@@

%\subsubsection{Semantic}

We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$. $\pi: I\rightarrow\mathbb{Z}$ is the interpretation of Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \pi\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution $(\eta, \pi)$ exists such as $\eta, \pi\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.
\begin{table}[h]
  \centering
  \begin{tabular}{lcl}
    $\eta,\pi \models \varphi_1\vee \varphi_2$ & $\mathsf{iff}$ & $\eta,\pi \models \varphi_1 \text{ or } \eta,\pi \models \varphi_2$ \\
    $\eta,\pi \models \neg\varphi $            & $\mathsf{iff}$ & $\eta,\pi \not\models \varphi$                                      \\
    $\eta, \pi \models x\in \regex$            & $\mathsf{iff}$ & $\exists w \in \lan(\regex),\eta,\pi \models x = w$                 \\
    $\eta, \pi \models \alpha_1 \leq \alpha_2$ & $\mathsf{iff}$ & $\pi(\alpha_1) \leq \pi(\alpha_2) $                                 \\
  \end{tabular}
  \caption{Semantics}
  \label{tab:semantics}
\end{table}
In addition to the classic regex operators (union, concatenation, closure), we syntactically support intersection, complement, and repetition. As we all know, the classic regular language is closed under intersection and complement \cite{aut_hopcraft}. Furthermore, the operation \emph{repetition} $\regex\{m,n\}$ means repeating regex $\regex$ at least $m$ times and at most $n$ times. It can be syntactically rewritten by concatenation and union:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $\regex$ $k$ times ($m\leq k\leq n$). So the regular language $\lan(\regex)$ defined in Table \ref{tab:syntax} is semantically equal to the classic regular language.
